---
title: "hcpd_regression"
author: "Briana Macedo"
date: '2024-09-05'
output: html_document
output_file: "zero-centered-smooths.html" 
output_dir: "~/Desktop/Main/macedo_tract/macedo_tracometry-main/plots/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
setwd("~/Desktop/Main/macedo_tract/macedo_tracometry-main/code/")
source("gam_visualize_model.R")
```

### Load packages
```{r}
pkgs <- c("mgcv",  "gamm4", "tidyverse", "readxl", "mgcViz", "DHARMa", "gratia",
  "ggforce", "marginaleffects", "dplyr", "ggplot2", "broom", "tidyr", "purrr", 
  "cowplot", "patchwork", "plotly")
invisible(lapply(pkgs, library, character.only = TRUE))
library(knitr)
```
### Read in the data
```{r}
metrics = read.csv("~/Desktop/Main/macedo_tract/outputs/metrics_df.csv")
main_metrics <- c('ad', 'area_of_end_region_1_mm2', 'area_of_end_region_2_mm2',
       'branch_volume_mm3', 'curl', 'diameter_mm', 'dti_fa', 'elongation',
       'gfa', 'ha', 'irregularity', 'irregularity_of_end_region_1',
       'irregularity_of_end_region_2', 'md', 'mean_length_mm', 'qa',
       'radius_of_end_region_1_mm', 'radius_of_end_region_2_mm', 'rd',
       'rdi', 'span_mm',
       'total_radius_of_end_regions_mm', 'total_surface_area_mm2',
       'trunk_volume_mm3',
       'volume_mm3')

tract_dict = read.csv("~/Desktop/Main/macedo_tract/tract_types_2.csv")

metrics <- metrics %>%
  left_join(tract_dict, by = c("bundle" = "Tract.Name")) %>%
  rename(tract_type = Type) 
```

```{r}
bundles = data.frame(bundle = unique(metrics$bundle))
bundles <- bundles %>%
  left_join(tract_dict, by = c("bundle" = "Tract.Name")) %>%
  rename(tract_type = Type) 
```

### Clean the data
```{r}
cleaned_metrics <- metrics %>%
  drop_na(score, sex, interview_age, t1_neighbor_corr, brain_size) %>%
  as.data.frame()

cleaned_metrics$brain_size <- as.numeric(scale(cleaned_metrics$brain_size))
cleaned_metrics$t1_neighbor_corr <- as.numeric(scale(cleaned_metrics$t1_neighbor_corr))
cleaned_metrics$sex <- as.factor(cleaned_metrics$sex)
cleaned_metrics$age = cleaned_metrics$interview_age/12
```

### Color by shape of curve
```{r}
color_by_shape <- function(modobj){
  age_p = summary(modobj)$s.table["s(age)", "p-value"] # TODO: multiple test-correction for p-values?
  if(age_p > 0.05) {
    return("light gray") # do not plot if age term is not significant
  }

  derv<-derivatives(modobj, select = "s(age)")
  derv<- derv %>%
      mutate(sig = !(0 > .lower_ci & 0 < .upper_ci))
  derv$sig_deriv = derv$.derivative*derv$sig # only plot deriv on the color bar if it is significant
  
  # if there is no significant change in derivative 
  if (min(derv$sig_deriv) == 0 & max(derv$sig_deriv) == 0) {
    return("light gray")
  }
  
  if (min(derv$.derivative) > 0) {
     return("red") # if always increasing
     } else if (min(derv$.derivative) < 0 & max(derv$.derivative) < 0) {
       return("blue") # if always decreasing
       } else {
         return("purple")
       }
}
```

```{r}
color_by_type <- function(modobj, tract_type){
  age_p = summary(modobj)$s.table["s(age)", "p-value"] # TODO: multiple test-correction for p-values?
  if(age_p > 0.05) {
    return("light gray") # do not plot if age term is not significant
  }

  if(tract_type[1] == "Association") { return("red")}
  if(tract_type[1] =="Commissure") {return("blue")}
  return("green") # projection
}
```

### Calculate partial R^2 using deviance
```{r}
calculate_partial_R2 <- function(modobj, data, covariates) {
  reduced_model <- gam(score ~ sex + t1_neighbor_corr, data = data, method = 'REML')
  D_full <- deviance(modobj)
  D_reduced <- deviance(reduced_model)
  partial_r2 <- (D_reduced - D_full) / D_reduced
  return(as.numeric(partial_r2, 3))
}
```

### Assign thicker lines to higher partial_r^2
```{r}
r2_to_thickness <- function(partial_r2) {
  line_sizes <- partial_r2 %>%
  mutate(thickness = case_when(
    r2 < 0.02     ~ 0.25,
    r2 < 0.09     ~ 1,
    r2 < 0.25     ~ 2.5,
    TRUE          ~ 5  # r2 >= 0.25
  ))
  return(line_sizes)
}
```

### GAM model - zero-centered smooths
Color:
<!-- red: increasing slope (significant derivative) -->
<!-- blue: decreasing slope (significant derivative) -->
<!-- purple: + and - slope (significant derivative) -->
<!-- gray: p-value for age term not significant or derivative's confidence interval contains 0 at all points -->
Association - red
Commissure - blue
Projection - green

Line width:
partial r2 < 0.02     ~ 0.25,
partial r2 < 0.09     ~ 1,
partial r2 < 0.25     ~ 2.5,
partial >= 0.25         ~ 3  # r2 >= 0.25

```{r}
measure = "score"
smooth_var = "age"
covariates = "sex + t1_neighbor_corr"
knots = 3
font_size = 20
theme_set(theme_classic(base_family = "sans", base_size = font_size))
line_size = 1.5
point_size = 2
bundles = unique(cleaned_metrics$bundle)
plot_list <- list()
counter = 0 

smooth_fits_all <- list()
all_partial_r2 <- data.frame(metric = character(), bundle = character(), r2 = numeric())

for (met in main_metrics) {
  counter = counter + 1
  metrics_span <- cleaned_metrics %>%
    filter(metric == met)
  
  model_formula <- as.formula(sprintf("%s ~ s(%s, k=knots, fx = T) + %s", measure, smooth_var, covariates))
  
  colors <- metrics_span %>%
    group_by(bundle) %>%
    do(color = color_by_shape(gam(model_formula, data = ., method = 'REML')))
  
  partial_r2 <- metrics_span %>%
    group_by(bundle) %>%
    do(r2 = calculate_partial_R2(gam(model_formula, data = ., method = 'REML'), data = ., covariates = covariates))
  
  partial_r2 = r2_to_thickness(partial_r2)
  partial_r2$r2 <- as.numeric(partial_r2$r2)
  partial_r2$metric <- met
  
  # ✅ Collect all R² values
  all_partial_r2 <- bind_rows(all_partial_r2, partial_r2)

  smooth_fits <- metrics_span %>% 
    group_by(bundle) %>% 
    do(fit = smooth_estimates(gam(model_formula, data = ., method = 'REML'), n = 100)) %>% 
    unnest(cols = fit) %>%
    left_join(colors, by = "bundle")  %>%
    left_join(partial_r2, by = "bundle") %>%
    mutate(color = as.character(color))
  
  smooth_fits <- smooth_fits %>%
    left_join(tract_dict, by = c("bundle" = "Tract.Name")) %>%
    rename(tract_type = Type) %>%
    mutate(metric = met)
  
  smooth_fits_all[[counter]] = smooth_fits
  
  print(plot_zero.centered_smooths(smooth_fits, ylab = met, thickness = smooth_fits$thickness))
}

# Rank metrics by mean R² and print
ranked_metrics <- all_partial_r2 %>%
  group_by(metric) %>%
  summarise(mean_r2 = mean(r2, na.rm = TRUE)) %>%
  arrange(desc(mean_r2))

cat("Metrics ranked by mean Partial R² (age-effect):\n")
print(ranked_metrics)

# Optional: Visualize R² ranking
ggplot(ranked_metrics, aes(x = reorder(metric, mean_r2), y = mean_r2)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Metrics Ranked by Age-Effect (Partial R²)", x = "Metric", y = "Mean Partial R²") +
  theme_minimal(base_size = font_size)

```
```{r fig.width=15, fig.height=12, dpi=300}
ggplot(ranked_metrics, aes(x = reorder(metric, mean_r2), y = mean_r2)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(
    title = "Metrics Ranked by Age-Effect (Partial R²)", 
    x = "Metric", 
    y = "Mean Partial R²"
  ) +
  theme_minimal(base_size = font_size + 4) +  # Increase base font size
  theme(
    plot.title = element_text(size = font_size + 6, face = "bold"),
    axis.title = element_text(size = font_size + 4),
    axis.text.y = element_text(size = font_size + 2),
    axis.text.x = element_text(size = font_size + 2),
    plot.margin = margin(20, 20, 20, 20),  # Adjust margins around the plot
    panel.grid.major.y = element_blank()   # Remove horizontal grid lines for clarity
  )
```
### Age effects of each metric per bundle
```{r fig.width=15, fig.height=12, dpi=300}

measure = "score"
smooth_var = "age"
covariates = "sex + t1_neighbor_corr"
knots = 3
significance_threshold <- 0.05  # P-value threshold for significance

font_size = 20
theme_set(theme_classic(base_family = "sans", base_size = font_size))

heatmap_data <- data.frame(metric = character(), bundle = character(), r2 = numeric(), p_value = numeric())

end_region_metrics <- c('area_of_end_region_1_mm2', 'area_of_end_region_2_mm2',
                        'irregularity_of_end_region_1',
       'irregularity_of_end_region_2', 
       'radius_of_end_region_1_mm', 'radius_of_end_region_2_mm', 
       'total_radius_of_end_regions_mm')

# Loop through each metric and bundle to compute R² and p-value
for (met in end_region_metrics) {
  metrics_span <- cleaned_metrics %>%
    filter(metric == met)
  
  for (b in unique(metrics_span$bundle)) {
    bundle_data <- metrics_span %>%
      filter(bundle == b)
    
    if (nrow(bundle_data) > 0) {
      model_formula <- as.formula(sprintf("%s ~ s(%s, k=%d, fx=T) + %s", measure, smooth_var, knots, covariates))
      gam_model <- gam(model_formula, data = bundle_data, method = 'REML')
      
      # Extract partial R²
      partial_r2 <- calculate_partial_R2(gam_model, data = bundle_data, covariates = covariates)
      
      # Extract p-value for age term
      age_p <- summary(gam_model)$s.table["s(age)", "p-value"]
      
      # Store results
      heatmap_data <- rbind(
        heatmap_data,
        data.frame(metric = met, bundle = b, r2 = as.numeric(partial_r2), p_value = age_p)
      )
    }
  }
}

# Mask R² values where p-value is not significant
heatmap_data <- heatmap_data %>%
  mutate(r2_masked = ifelse(p_value < significance_threshold, r2, NA))

```


```{r fig.width=20, fig.height=25, dpi=300}
# Create heatmap with bundles on Y-axis and extended height
ggplot(heatmap_data, aes(x = metric, y = bundle, fill = r2_masked)) +
  geom_tile(color = "white") +  # White gridlines
  scale_fill_gradient(
    low = "white", high = "steelblue", na.value = "white",  # White for non-significant
    name = "Partial R²"
  ) +
  labs(
    title = "Heatmap of Age-Effect (Partial R²) per Bundle and Metric",
    x = "Metric",
    y = "Bundle"
  ) +
  theme_minimal(base_size = font_size) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),  # Rotate X-axis labels for clarity
    axis.text.y = element_text(size = font_size),                    # Keep Y-axis readable
    plot.title = element_text(hjust = 0.5, size = font_size + 4),
    legend.title = element_text(size = font_size + 2),
    legend.text = element_text(size = font_size),
    plot.margin = margin(30, 30, 30, 30)                             # Extra space around the plot
  )

ggsave("ageeffects_plot.png", width = 20, height = 25, dpi = 300)
```

### Male vs female models and compatison
```{r}
measure = "score"
smooth_var = "age"
knots = 3
font_size = 20
theme_set(theme_classic(base_family = "sans", base_size = font_size))
significance_threshold <- 0.05  # P-value threshold for significance'

cleaned_metrics$sex <- factor(cleaned_metrics$sex, levels = c(0, 1))
all_partial_r2_sex <- data.frame(metric = character(), bundle = character(), sex = character(), r2 = numeric())
heatmap_data_sex <- data.frame(metric = character(), bundle = character(), sex = character(), r2 = numeric(), p_value = numeric())

for (met in main_metrics) {
  metrics_span <- cleaned_metrics %>%
    filter(metric == met)

  for (b in unique(metrics_span$bundle)) {
    for (gender in c(0, 1)) {  # Assuming 0 = Female, 1 = Male
      bundle_data <- metrics_span %>%
        filter(bundle == b, sex == gender)
      
      # Convert 'sex' to numeric to avoid factor contrast issues
      bundle_data$sex <- as.numeric(as.character(bundle_data$sex))

      if (nrow(bundle_data) > 0) {
        model_formula <- as.formula(sprintf("%s ~ s(%s, k=%d, fx=T) + t1_neighbor_corr",
                                            measure, smooth_var, knots))
        gam_model <- gam(model_formula, data = bundle_data, method = 'REML')

        # Extract partial R² and p-value
        partial_r2 <- calculate_partial_R2(gam_model, data = bundle_data, covariates = "t1_neighbor_corr")
        age_p <- summary(gam_model)$s.table["s(age)", "p-value"]

        # Store results
        all_partial_r2_sex <- rbind(
          all_partial_r2_sex,
          data.frame(metric = met, bundle = b, sex = gender, r2 = as.numeric(partial_r2))
        )
      }
    }
  }
}


print(head(all_partial_r2_sex))

# Mask non-significant R² values
heatmap_data_sex <- heatmap_data_sex %>%
  mutate(r2_masked = ifelse(p_value < significance_threshold, r2, NA))

# Recode 'sex' column from 1/0 to "Male"/"Female"
all_partial_r2_sex <- all_partial_r2_sex %>%
  mutate(sex = ifelse(sex == 1, "Male", "Female"))

# Compare Average Age Effects (Male vs. Female)
avg_age_effects <- all_partial_r2_sex %>%
  group_by(metric, sex) %>%
  summarise(mean_r2 = mean(r2, na.rm = TRUE), .groups = 'drop') %>%
  pivot_wider(names_from = sex, values_from = mean_r2) %>%
  mutate(difference = Male - Female)  # 💡 Now should work without error

# Plot the comparison
ggplot(avg_age_effects, aes(x = Male, y = Female, label = metric)) +
  geom_point(color = "darkgreen", size = 3) +
  geom_text(vjust = -0.5, size = 4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Comparison of Average Age Effects (Partial R²) in Male vs. Female Models",
    x = "Male Model (Mean Partial R²)",
    y = "Female Model (Mean Partial R²)"
  ) +
  theme_minimal(base_size = font_size) +
  theme(plot.title = element_text(size = font_size + 4, face = "bold"))
```

```{r fig.width=20, fig.height=25, dpi=300}
# 💡 Heatmap: Male vs. Female
heatmap_data_sex <- all_partial_r2_sex %>%
  pivot_wider(names_from = sex, values_from = r2) %>%
  mutate(difference = Male - Female)

ggplot(heatmap_data_sex, aes(x = metric, y = bundle, fill = difference)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name = "Partial R² Diff") +
  labs(title = "Heatmap of Age-Effect Difference (Male vs. Female)", x = "Metric", y = "Bundle") +
  theme_minimal(base_size = font_size) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

```

```{r fig.width=20, fig.height=25, dpi=300}
cleaned_metrics$sex <- as.numeric(as.character(cleaned_metrics$sex))

# 🔍 Metrics of interest
end_region_metrics <- c('area_of_end_region_1_mm2', 'area_of_end_region_2_mm2',
                        'irregularity_of_end_region_1', 'irregularity_of_end_region_2', 
                        'radius_of_end_region_1_mm', 'radius_of_end_region_2_mm', 
                        'total_radius_of_end_regions_mm')

measure = "score"
smooth_var = "age"
covariates = "t1_neighbor_corr"  # Exclude 'sex' for sex-stratified analysis
knots = 3
font_size = 16
theme_set(theme_classic(base_family = "sans", base_size = font_size))
significance_threshold <- 0.05

# Initialize data frames
heatmap_data_sex <- data.frame(metric = character(), bundle = character(), sex = character(), r2 = numeric(), p_value = numeric())

# Loop for each metric and sex
for (met in end_region_metrics) {
  metrics_span <- cleaned_metrics %>%
    filter(metric == met)
  
  for (gender in c(0, 1)) {
    sex_label <- ifelse(gender == 1, "Male", "Female")
    metrics_sex <- metrics_span %>% filter(sex == gender)
    
    for (b in unique(metrics_sex$bundle)) {
      bundle_data <- metrics_sex %>%
        filter(bundle == b)
      
      if (nrow(bundle_data) > 1) {  # ⚡ Ensure sufficient data points
        model_formula <- as.formula(sprintf("%s ~ s(%s, k=%d, fx=T) + %s", measure, smooth_var, knots, covariates))
        gam_model <- gam(model_formula, data = bundle_data, method = 'REML')
        
        # 🎯 Extract partial R²
        partial_r2 <- calculate_partial_R2(gam_model, data = bundle_data, covariates = covariates)
        age_p <- summary(gam_model)$s.table["s(age)", "p-value"]
        
        # 💾 Store results
        heatmap_data_sex <- rbind(
          heatmap_data_sex,
          data.frame(metric = met, bundle = b, sex = sex_label, r2 = as.numeric(partial_r2), p_value = age_p)
        )
      }
    }
  }
}

# Mask non-significant R² values
heatmap_data_sex <- heatmap_data_sex %>%
  mutate(r2_masked = ifelse(p_value < significance_threshold, r2, NA))

# Common scale across plots
max_r2 <- max(heatmap_data_sex$r2_masked, na.rm = TRUE)

# Heatmap for Males
heatmap_male <- heatmap_data_sex %>%
  filter(sex == "Male") %>%
  ggplot(aes(x = metric, y = bundle, fill = r2_masked)) +
  geom_tile(color = "white") +
  scale_fill_gradient(
    low = "white", high = "steelblue", na.value = "white", 
    limits = c(0, max_r2), name = "Partial R²"
  ) +
  labs(title = "Age-Effect Heatmap (Male)", x = "Metric", y = "Bundle") +
  theme_minimal(base_size = font_size) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Heatmap for Females
heatmap_female <- heatmap_data_sex %>%
  filter(sex == "Female") %>%
  ggplot(aes(x = metric, y = bundle, fill = r2_masked)) +
  geom_tile(color = "white") +
  scale_fill_gradient(
    low = "white", high = "steelblue", na.value = "white", 
    limits = c(0, max_r2), name = "Partial R²"
  ) +
  labs(title = "Age-Effect Heatmap (Female)", x = "Metric", y = "Bundle") +
  theme_minimal(base_size = font_size) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Combine Male and Female heatmaps side-by-side
(heatmap_male | heatmap_female) +
  plot_annotation(title = "Comparison of Age Effects by Sex (Selected End Region Metrics)",
                  theme = theme(plot.title = element_text(size = font_size + 4, hjust = 0.5, face = "bold")))
ggsave("ageeffects_by_sex_plot.png", width = 20, height = 25, dpi = 300)
```
```{r fig.width=20, fig.height=25, dpi=300}
cleaned_metrics$sex <- ifelse(cleaned_metrics$sex == 1, "Male", "Female")

# Loop to calculate mean partial R² by metric and sex
mean_r2_by_sex <- all_partial_r2_sex %>%
  group_by(metric, sex) %>%
  summarise(mean_r2 = mean(r2, na.rm = TRUE), .groups = "drop")

# Ensure common y-axis scale across both plots
max_r2 <- max(mean_r2_by_sex$mean_r2, na.rm = TRUE)

# Plot for Males
male_plot <- mean_r2_by_sex %>%
  filter(sex == "Male") %>%
  ggplot(aes(x = reorder(metric, mean_r2), y = mean_r2)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  ylim(0, max_r2) +  # ⚡ Common scale
  labs(
    title = "Mean Age-Effect (Partial R²) - Male",
    x = "Metric",
    y = "Mean Partial R²"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16)
  )

# Plot for Females
female_plot <- mean_r2_by_sex %>%
  filter(sex == "Female") %>%
  ggplot(aes(x = reorder(metric, mean_r2), y = mean_r2)) +
  geom_bar(stat = "identity", fill = "lightpink") +
  coord_flip() +
  ylim(0, max_r2) +  # ⚡ Common scale
  labs(
    title = "Mean Age-Effect (Partial R²) - Female",
    x = "Metric",
    y = "Mean Partial R²"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16)
  )

# Combine plots side by side
(male_plot | female_plot) +
  plot_annotation(
    title = "Comparison of Mean Age-Effects by Sex (Across All Bundles)",
    theme = theme(plot.title = element_text(size = 20, hjust = 0.5, face = "bold"))
  )
```

```{r}
smooth_fits_final = bind_rows(smooth_fits_all)
smooth_fits_final <- smooth_fits_final %>%
  mutate(across(where(is.list), ~sapply(., function(x) if (length(x) == 1) x else NA)))
write.csv(smooth_fits_final, "~/Desktop/Main/macedo_tract/outputs/smooth_fits_100.csv", row.names = FALSE)
```

### SLF and TR

```{r fig.width=25, fig.height=15, dpi=300}
library(ggrepel)
library(patchwork)  # For combining plots

measure = "score"
smooth_var = "age"
covariates = "sex + t1_neighbor_corr"
knots = 3
font_size = 20
theme_set(theme_classic(base_family = "sans", base_size = font_size))
line_size = 1.5
point_size = 2
highlight_bundles <- c("SuperiorLongitudinalFasciculus1L", "SuperiorLongitudinalFasciculus1R",
                       "FornixR", "FornixL", "ThalamicRadiationAnteriorL", "ThalamicRadiationAnteriorR")

assign_bundle_colors <- function(bundle) {
  colors_map <- c(
    "SuperiorLongitudinalFasciculus1L" = "red",
    "SuperiorLongitudinalFasciculus1R" = "blue",
    "FornixR" = "green",
    "FornixL" = "orange",
    "ThalamicRadiationAnteriorL" = "purple",
    "ThalamicRadiationAnteriorR" = "brown"
  )
  return(colors_map[bundle])
}

bundles = highlight_bundles
plot_list <- list()
counter = 0 

smooth_fits_all <- list()
all_partial_r2 <- data.frame(metric = character(), bundle = character(), r2 = numeric())

for (met in end_region_metrics) {
  counter = counter + 1
  metrics_span <- cleaned_metrics %>%
    filter(metric == met & bundle %in% highlight_bundles)

  model_formula <- as.formula(sprintf("%s ~ s(%s, k=knots, fx = T) + %s", measure, smooth_var, covariates))

  partial_r2 <- metrics_span %>%
    group_by(bundle) %>%
    do(r2 = calculate_partial_R2(gam(model_formula, data = ., method = 'REML'), data = ., covariates = covariates))

  partial_r2 = r2_to_thickness(partial_r2)
  partial_r2$r2 <- as.numeric(partial_r2$r2)
  partial_r2$metric <- met

  all_partial_r2 <- bind_rows(all_partial_r2, partial_r2)

  smooth_fits <- metrics_span %>% 
    group_by(bundle) %>% 
    do(fit = smooth_estimates(gam(model_formula, data = ., method = 'REML'), n = 100)) %>% 
    unnest(cols = fit) %>%
    left_join(partial_r2, by = "bundle") %>%
    mutate(color = assign_bundle_colors(bundle)) %>%
    left_join(tract_dict, by = c("bundle" = "Tract.Name")) %>%
    rename(tract_type = Type) %>%
    mutate(metric = met)

  smooth_fits_all[[counter]] = smooth_fits

  p <- ggplot(smooth_fits, aes(x = age, y = .estimate, color = bundle, group = bundle)) +
    geom_line(linewidth = line_size) +
    scale_color_manual(
      values = setNames(assign_bundle_colors(highlight_bundles), highlight_bundles),
      breaks = highlight_bundles,
      name = "Highlighted Bundles"
    ) +
    labs(y = met, x = "Age") +
    theme(legend.position = "right")

  plot_list[[counter]] <- p
}

# Combine all plots into one subplot
combined_plot <- wrap_plots(plotlist = plot_list, ncol = 2)
print(combined_plot)

ranked_metrics <- all_partial_r2 %>%
  group_by(metric) %>%
  summarise(mean_r2 = mean(r2, na.rm = TRUE)) %>%
  arrange(desc(mean_r2))

cat("Metrics ranked by mean Partial R² (age-effect):\n")
print(ranked_metrics)

ggplot(ranked_metrics, aes(x = reorder(metric, mean_r2), y = mean_r2)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Metrics Ranked by Age-Effect (Partial R²)", x = "Metric", y = "Mean Partial R²") +
  theme_minimal(base_size = font_size)

```

```

### PCA - sanity check, compared to PCA from python
```{r}
smooth_fits_small = smooth_fits_final %>%
  select(bundle, .estimate, age, metric)

metric_fits <- smooth_fits_small %>% 
  filter(smooth_fits_small$metric == "gfa") %>%
  select(-metric)

smooth_fits_wide <- metric_fits %>%
  pivot_wider(names_from = age, values_from = .estimate) %>%
  select(-bundle)

gfa_PCA <- prcomp(smooth_fits_wide, center = TRUE, scale = TRUE)
factoextra::fviz_eig(gfa_PCA)
pca_loadings <- as.data.frame(gfa_PCA$rotation[,1:3])
pca_loadings$age <- as.numeric(row.names(pca_loadings))
ggplot(pca_loadings, aes(age, PC1)) + geom_point() + labs(title = "PC1", x = "Predicted Age", y = "PC1 loading") + theme_minimal()
```

### GAM - separate plots, by metric
```{r, fig.height=40, fig.width=40, eval=FALSE}
measure = "score"
smooth_var = "age"
covariates = "sex + t1_neighbor_corr"
knots = 3
font_size = 14
theme_set(theme_classic(base_family = "sans", base_size = font_size))
line_size = 1.5
point_size = 2
bundles = unique(cleaned_metrics$bundle)

end_region_metrics <- c('area_of_end_region_1_mm2', 'area_of_end_region_2_mm2',
                        'irregularity_of_end_region_1', 'irregularity_of_end_region_2', 
                        'radius_of_end_region_1_mm', 'radius_of_end_region_2_mm', 
                        'total_radius_of_end_regions_mm')

for (met in mnd_region_metrics) {

  plots <- list()
  counter <- 0

  for (bund in bundles) {
    counter = counter + 1
    metrics_span <- cleaned_metrics %>%
    filter(metric == met, bundle == bund)

    model_formula <- as.formula(sprintf("%s ~ s(%s, k=knots, fx = T) + %s", measure, smooth_var, covariates))
    gam_model <- gam(model_formula, data = metrics_span, method = 'REML')
    
    p = visualize_model(modobj = gam_model, smooth_var = smooth_var, group_var = "sub", int_var="sex", custom_color="red", plabels=bund, derivative_plot = T, print_on = FALSE)

    plots[[counter]] = p
  }

  combined_plot <- plot_grid(plotlist = plots, ncol = 8)
  final_plot_with_title <- plot_grid(
    ggdraw() + draw_label(met, fontface = 'bold', size = 30),
    combined_plot, ncol = 1, rel_heights = c(0.1, 1))
  print(final_plot_with_title)
}
```

### GAM - separate plots, by bundle
```{r, fig.height=40, fig.width=40, eval=FALSE}
measure = "score"
smooth_var = "age"
covariates = "sex + t1_neighbor_corr"
knots = 3
font_size <- 14
theme_set(theme_classic(base_family = "sans", base_size = font_size))
line_size <- 1.5
point_size <- 2
bundles = unique(cleaned_metrics$bundle)
i


main_metrics <- c()

for (bund in bundles) {
  plots <- list()
  counter <- 0
  
  for (met in main_metrics ) {
    counter = counter + 1
    metrics_span <- cleaned_metrics %>%
    filter(metric == met, bundle == bund)
    
    model_formula <- as.formula(sprintf("%s ~ s(%s, k=knots, fx = T) + %s", measure, smooth_var, covariates)) 
    gam_model <- gam(model_formula, data = metrics_span, method = 'REML')
    # print(bund)
    # print(summary(gam_model))
    # print(anova(gam_model, test = "F", type = "R2"))
  
    p = visualize_model(modobj = gam_model, smooth_var = smooth_var, group_var = "sub", int_var="sex", custom_color="red", plabels=met, derivative_plot = T, print_on = FALSE)
    
    plots[[counter]] = p
    
  }
  
  combined_plot <- plot_grid(plotlist = plots, ncol = 5)
  final_plot_with_title <- plot_grid(
  ggdraw() + draw_label(bund, fontface = 'bold', size = 30),
  combined_plot, ncol = 1, rel_heights = c(0.1, 1))
  print(final_plot_with_title)
}

```


### PCA - sanity check, compared to PCA from python

```{r}
smooth_fits_small = smooth_fits_final %>%
  select(bundle, .estimate, age, metric)

metric_fits <- smooth_fits_small %>% 
  filter(smooth_fits_small$metric == "gfa") %>%
  select(-metric)

smooth_fits_wide <- metric_fits %>%
  pivot_wider(names_from = age, values_from = .estimate) %>%
  select(-bundle)

gfa_PCA <- prcomp(smooth_fits_wide, center = TRUE, scale = TRUE)
factoextra::fviz_eig(gfa_PCA)
pca_loadings <- as.data.frame(gfa_PCA$rotation[,1:3])
pca_loadings$age <- as.numeric(row.names(pca_loadings))
ggplot(pca_loadings, aes(age, PC1)) + geom_point() + labs(title = "PC1", x = "Predicted Age", y = "PC1 loading") + theme_minimal()
```

### GAM - separate plots, by metric
```{r, fig.height=40, fig.width=40, eval=FALSE}
measure = "score"
smooth_var = "age"
covariates = "sex + t1_neighbor_corr"
knots = 3
font_size = 14
theme_set(theme_classic(base_family = "sans", base_size = font_size))
line_size = 1.5
point_size = 2
bundles = unique(cleaned_metrics$bundle)

for (met in main_metrics) {

  plots <- list()
  counter <- 0

  for (bund in bundles) {
    counter = counter + 1
    metrics_span <- cleaned_metrics %>%
    filter(metric == met, bundle == bund)

    model_formula <- as.formula(sprintf("%s ~ s(%s, k=knots, fx = T) + %s", measure, smooth_var, covariates))
    gam_model <- gam(model_formula, data = metrics_span, method = 'REML')
    
    p = visualize_model(modobj = gam_model, smooth_var = smooth_var, group_var = "sub", int_var="sex", custom_color="red", plabels=bund, derivative_plot = T, print_on = FALSE)

    plots[[counter]] = p
  }

  combined_plot <- plot_grid(plotlist = plots, ncol = 8)
  final_plot_with_title <- plot_grid(
    ggdraw() + draw_label(met, fontface = 'bold', size = 30),
    combined_plot, ncol = 1, rel_heights = c(0.1, 1))
  print(final_plot_with_title)
}
```

### GAM - separate plots, by bundle
```{r, fig.height=40, fig.width=40, eval=FALSE}
measure = "score"
smooth_var = "age"
covariates = "sex + t1_neighbor_corr"
knots = 3
font_size <- 14
theme_set(theme_classic(base_family = "sans", base_size = font_size))
line_size <- 1.5
point_size <- 2
bundles = unique(cleaned_metrics$bundle)

main_metrics <- c()

for (bund in bundles) {
  plots <- list()
  counter <- 0
  
  for (met in main_metrics ) {
    counter = counter + 1
    metrics_span <- cleaned_metrics %>%
    filter(metric == met, bundle == bund)
    
    model_formula <- as.formula(sprintf("%s ~ s(%s, k=knots, fx = T) + %s", measure, smooth_var, covariates)) 
    gam_model <- gam(model_formula, data = metrics_span, method = 'REML')
    # print(bund)
    # print(summary(gam_model))
    # print(anova(gam_model, test = "F", type = "R2"))
  
    p = visualize_model(modobj = gam_model, smooth_var = smooth_var, group_var = "sub", int_var="sex", custom_color="red", plabels=met, derivative_plot = T, print_on = FALSE)
    
    plots[[counter]] = p
    
  }
  
  combined_plot <- plot_grid(plotlist = plots, ncol = 5)
  final_plot_with_title <- plot_grid(
  ggdraw() + draw_label(bund, fontface = 'bold', size = 30),
  combined_plot, ncol = 1, rel_heights = c(0.1, 1))
  print(final_plot_with_title)
}

```
  


